---
title: "Activity 8: Computational thinking 2: conditionals"
subtitle: Feb. 2nd, 2026, Calvin Munson
format: gfm
execute:
  warning: false
editor: source
---

# Overview

Welcome! This is the second "computational thinking" lesson, which continues work on for loops and revisits conditional statements.

------------------------------------------------------------------------

You will submit one output for this activity:

1.  A **PDF** of a rendered Quarto document with all of your R code. Please create a new Quarto document (e.g. don't use this `README.qmd`), include all of the code that appears in this document, in addition to adding your own code and **answers to all of the questions** in the "Q#" sections. Submit this through Gradescope.

*If you have trouble submitting as a PDF, please ask Calvin or Malin for help. If we still can't solve it, you can submit the .qmd file instead.*

A reminder: **Please label the code** in your final submission in two ways: 1) denote your answers to each question using headers that correspond to the question you're answering and 2) thoroughly "comment" your code: remember, this means annotating your code directly by typing descriptions of what each line does after a `#`. This will help future you!

------------------------------------------------------------------------

Let's start by reading in the relevant packages

```{r}
library(tidyverse)
library(here)
```

------------------------------------------------------------------------

# 1. Conditionals

We've already worked with conditionals during one of the seaside chats, which used `case_when()`, `dplyr's` version of an `if_else` statement. Today we'll walk through using `if_else` statements, the precursor to `case_when()`, as well as get more practice with `case_when()`.

## 1.1 if else statements

`if else` statements take the form of:

``` default
if (condition){
  code to execute if condition is TRUE
} else {
  code to execute if condition is FALSE
}
```

For example, below we define x as 5, then write a statement that checks whether or not a statement, `x > 10`, is TRUE. If it's true, it will print out the first part. If not, it will print the second.

```{r}
x <- 5
 
# Check value is less than or greater than 10 
if(x > 10)
{ 
    print(paste(x, "is greater than 10")) 
} else
{ 
    print(paste(x, "is less than or equal to 10")) 
}
```


### Q1.1: Modify the value of x

Copy/paste this example by change x to be 11. What is the output of the if else statement?


```{r}
x <- 11
 
# Check if the value of x is less than or equal OR greater than 10 
if(x > 10)
{ 
    print(paste(x, "is greater than 10")) 
} else
{ 
    print(paste(x, "is less than or equal to 10")) 
}

```


---------

Let's make it a little more complex by adding a third conditional. The if else statements in R can be nested together to form multiple statements and will evaluate expressions based on the conditions one by one, beginning from the outer condition to the inner one.

Here, we will make R print one thing if x is less than 10, something else if x is equal to 10, and a third thing if x is greater than 10.

```{r}
# define a variable
x <- 10

# check the value of x using nested if-else statements
if (x < 10) {
  # if x is less than 10
  print("x is less than 10")
} else {
  # if x is exactly equal to 10
  if (x == 10) {
    # if x equals 10
    print("x is 10!!!")
  } else {
    # if x is greater than 10
    print("x is greater than 10")
  }
}

```

Run the statement above with `x = 9`, `x = 10`, and `x = 11` to see what happens.

--------

Let's combine what we learned about for loops with an if else statement. The goal here is to take a vector of numbers, loop through it, and apply our if else statement to each element in the vector, telling us if it's less than, equal to, or greater than 10.

First define a vector to loop through:

```{r}
vec <- c(10, 11, 9, 10)
```

Then we start the for loop:

```{r}
# For 1 through the length of the vector "vec" 
for (i in 1:length(vec)) {
  
# check the value of using nested if-else statements
if (vec[i] < 10) {
  # if the element is less than 10
  print("value is less than 10")
} else {
  # if the element is exactly equal to 10
  if (vec[i] == 10) {
    # if the element equals 10
    print("value is 10!!!")
  } else {
    # if the element is greater than 10
    print("value is greater than 10")
  }
}

}

```

The for loops iterates through the vector and applies the if else statement separately for each vector element, printing an output for each vector element.

### Q1.2: Create a new for loop + if else statement

Remake a similar for loop / if else statement as above. The for loop should loop through a vector called `y` and should assess whether or not the values are negative, positive, or equal to zero.
First define a vector to loop through:

```{r}
# Define the vector
y <- c(-2, 42, 0, 10)

# For 1 through the length of the vector "y"
for (i in 1:length(y)) {
  
# check the value of using nested if-else statements
if (y[i] < 0) {
  # if the element is less than 0
  print("value is less than 0")
} else {
  # if the element is exactly equal to 0
  if (y[i] == 0) {
    # if the element equals 0
    print("value is 0!!!")
  } else {
    # if the element is greater than 0
    print("value is greater than 0")
  }
}

}

```

---------

## 1.2 `case_when()`

------------

# 2. Getting help in R

Early in your R journey you will likely have errors in your R code more often than not, so knowing how to deal with code that doesn't work as you expect is a key part of learning. You have already been asked in past lessons to deal with this, by identifying error messages, accessing help pages, and going online to .

------

## 2.1 Help pages

-------

## 2.2 Stack overflow

-------

## 2.3 Traceback()?

-----


## 2.4 LLMs

-------





